---
title: "Reproductive Search - Data Science"
author: "DBW"
date: "9 de julho de 2015"
output: html_document
---
```{r, echo=TRUE}
library(dplyr)
library(ggplot2)

```

##Loading and checking the data with the following code:
checking the data frame
```{r, echo=TRUE}
dataframe<-read.csv("activity.csv")
str(dataframe)
head(dataframe)
tail(dataframe)
```

##What is the mean total number of steps taken per day?

###for this part, we can ignore the 'NAs', so we will use the whole dataframe
First we will group the dataframe by date, then we will calculate the total steps by group (date)
```{r, echo=TRUE}
#grouping the data by day and calculating the total steps per day
grouped<-group_by(dataframe,date)
total<-summarize(grouped, total_steps = sum(steps, na.rm=TRUE))


#making a histogram of the total steps
hist(total$total_steps, main = "Total steps frequency", xlab="total steps", col="blue", breaks=15)

#Calculate and report the mean and median of the total numer of steps taken per day

#mean of the total steps by day
mean(total$total_steps)

#median of the total steps by day
median(total$total_steps)
```

##What is the average daily activity pattern?
First we should group the dataframes by interval, then calculate the average of the groups (the intervals).
```{r, echo=TRUE}
#calculating the average of steps in each interval, by day
intervalmean<-group_by(dataframe,interval)
intervalmean<-summarize(intervalmean, interval_mean = mean(steps, na.rm=TRUE))



#making the plot
with(intervalmean, plot(interval, interval_mean,main= "Average of steps by interval of the day", xlab="time", ylab = "steps average", type = "l",col="blue"))

#finding wich time interval has the maximum average number of steps
subset(intervalmean, 
       intervalmean$interval_mean==max(intervalmean$interval_mean))[,"interval"]
```

##Imputing missing values
###we want the number of rows which have NA value in any of the columns of the dataframe
first we should check for NA in each column and then subset the dataframe for the columns which have NA
```{r, echo=TRUE}
#first we check for NA in each column
for(i in 1:ncol(dataframe)){
        print(any(is.na(dataframe[,i])))
}

#now we know only the first column has NAs
subset(dataframe ,is.na(dataframe[,1]))%>%
nrow
```

#Devising a strategy to fill the missing values
We will use the corrected average of steps by interval:
the correction is the product of the average of this interval by the percentage of the mean of that day compared with the median (50% of the values) of all the means.
```{r, echo=TRUE}
#calculating the mean of steps by day
daymean<-group_by(dataframe,date)
daymean<-summarize(daymean, day_mean = mean(steps, na.rm=TRUE))
head(daymean)

#calculating the percentage of the mean of each day compared to the median of means of all day
daymean<-mutate(daymean, correction = day_mean/median(day_mean, na.rm=TRUE))
head(daymean, 20)

#modyfing the dataframe to have the interval mean by each interval
newdf<-left_join(dataframe, intervalmean, by="interval")
head(newdf, 20)

#modyfing the dataframe to have the correction by each day
correction<-daymean[,c("date","correction")]
newdf<-left_join(newdf, correction, by="date")
set.seed(1000)
sample_n(newdf, 30)

```

###Some aesthethics
We will assume that if the correction is NaN (because there was no mean for this day), we should not correct it (corrrection = 1)
```{r, echo=TRUE}
#formatting the interval
#newdf$interval<-formatC(intervalmean$interval/100,digits =2,
                               #width =5, format="f", flag="0")
#removing NaNs from correction
newdf<-mutate(newdf, correction = ifelse(is.na(correction),1,correction))
set.seed(1000)
sample_n(newdf, 30)
```

###replacting NA steps by the corrected interval mean as explained above

```{r, echo=TRUE}
#if steps==NA it will make the mutation, if not it will maintain the value of steps
newdf<-mutate(newdf, steps = ifelse(is.na(steps),interval_mean*correction,steps))
head(newdf)
set.seed(1000)
sample_n(newdf, 30)
```

###calculating the total steps by interval, also the mean and the median of the total steps

```{r, echo=TRUE}
#grouping the data by day and calculating the total steps per day
gr_new<-group_by(newdf,date)
total_new<-summarize(gr_new, total_steps = sum(steps))
head(total_new,20)


#making a histogram of the total steps
par(mfrow=c(1,1))
hist(total_new$total_steps, main = "Total steps frequency", xlab="total steps", col="blue", breaks=15)

#Calculate and report the mean and median of the total numer of steps taken per day

#mean of the total steps by day
mean(total_new$total_steps)

#median of the total steps by day
median(total_new$total_steps)
```
###Comparing values with NA and without NA

```{r}
#means
paste("Clean Data Mean:",round(mean(total_new$total_steps), digits = 2),"vs raw data (with NA) Mean:", round(mean(total$total_steps),digits=2))

#medians
paste("Clean Data Median:",round(median(total_new$total_steps),digits = 2),"vs raw data (with NA) Median:", round(median(total$total_steps),digits = 2))

#making 2 histograms, with both medians marked
par(mfrow=c(2,1),mar= c(4,1,1,1), oma=c(1,1,1,1))

hist(total_new$total_steps, main = "Total steps frequency Cleaned", xlab="total steps", col="blue", breaks=15)
abline(v=median(total_new$total_steps), col="green", lwd="4")
abline(v=median(total$total_steps), col="magenta", lwd="4")
legend("topright", lty=1, legend = c("clean median", "raw median"), col=c("green","magenta"), bty="n")

hist(total$total_steps, main = "Total steps frequency Raw (with NA)", xlab="total steps", col="blue", breaks=15)
abline(v=median(total_new$total_steps), col="green", lwd="4")
abline(v=median(total$total_steps), col="magenta", lwd="4")
legend("topright", lty=1, legend = c("clean median", "raw median"), col=c("green","magenta"), bty="n")
```

##Differences in activity patterns between weekdays and weekends:
we will first need to convert the dates to POSIXt format, than add the weekdays to the dataframe and than transform those weekdays to a new variable which differenatiaties between weekdays and weekends

```{r, echo=TRUE}
#formating date to POSIXt
newdf$date<-strptime(newdf$date, format = "%Y-%m-%d")
class(newdf$date)

#adding weekdays
week_data<-mutate(newdf, weekday = weekdays(newdf$date, abbreviate = TRUE))
week_data<-select(week_data, -interval_mean, -correction)
head(week_data)

#converting to weekend/weekday
week_data<-mutate(week_data, 
                  week_part = ifelse(weekday=="Sáb"|weekday=="Dom", 
                                     "weekend", "weekday"))
head(week_data)

#converting week_part to factor
week_data$week_part<-as.factor(week_data$week_part)
str(week_data)
```

##Making two panel time series of the 5-minutes interval and the average number of steps take, averaged across all weeday or weekend days

we will new to group the data by interval and then by week part to plot the panels

```{r, echo=TRUE}
#putting the date back to character to use the group_by function
week_data$date<-as.character(week_data$date)

#grouping by interval
grouped_week<-group_by(week_data, interval, week_part)
avg_week<-summarize(grouped_week, steps = mean(steps))
head(avg_week)

#making the plot
ggplot(avg_week, aes (interval, steps))+facet_wrap(~week_part, ncol=1)+geom_line(aes(group=week_part), color = "blue")+ggtitle("Average of steps in the weekdays vs weekends")

```

